Chemical tagging can work: \\ Identification of stellar phase-space structures \\ purely by chemical-abundance similarity

### 2016-01-19 papers/clusters.tex
abstract tweak
wording tweaks

\begin{abstract}
% Context:
Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric abundances.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

We have demonstrated for the first time that, in a large-scale
spectroscopic survey of stars, stars that appear similar in
chemical-abundance properties are often associated physically in
clusters or other structures.
The reverse is well established---that is, that stars that are
associated physically in clusters are similar in chemical-abundance
properties.
That is, although they often show chemical diversity, that chemical
diversity is small, and follows well-defined trends or a subspace in
abundance space (for example, \citealt{gratton, meszaros, bovy},
and references within those).
However, it was not known---prior to this work---whether abundance-space information would be
informative enough, or \emph{measurable} precisely enough to find this
structure in surveys dominated by a large mix of stars from different
origins and ages (and there was good reason to be pessimistic;
\citealt{ting}).
We have demonstrated with a straightforward (and in no way optimal)
clustering in chemical-abundance space that it is easy to find co-eval
stellar structures.

What made this possible is substantial improvements in precision
delivered by \thecannon.
We don't have an absolute measure of the precision of the measurements,
but from looking at open and globular clusters, it appears to be on the
order of or better than 0.04~dex for the median element in the list of 15.
These improvements are discussed in detail in the companion papers
\citep{casey16, ness16} about the abundance measurements.
We believe these improvements come from a combination of factors, not
limited to (a)~improvements in the determination of pseudo-continuum,
(b)~the use of more spectral range than just unblended element windows
\citep{aspcap}, and (c)~accurate spectral predictions
(\thecannon\ delivers accurate predictions because it is fit to
observed data). 
Furthermore, because abundance space is high in dimensionality, even
small improvements in abundance measurement get taken to a significant
power when thinking about information gains.

@@

We freely admit that k-means is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
highly similar abundances in 15 dimensions that are now spread throughout
the Galaxy. Because some stars are candidates of the massive globular
cluster M53, it provides tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo by a single globular
cluster early in the Milky Way's formation. While speculative (and we
make no consideration to whether this is indeed the true scenario),
this example demonstrates the ultimate goal of chemical tagging.

The abundance-space group of stars in the disk structure shown in
\figurename~\ref{fig:disk} may be associated with a very thin
Milky-Way bar component reported previously \citep{wegg}, which is
found to exist predominantly toward the end of the bar (at $l$
$\approx$ 30$^\circ$).
The stars in this structure are metal-rich and have a low
[C/N] ratio and are therefore likely young (following the [C/N]--mass
relationship; \citealt{martig}).
It is possible that this structure has very low scale height because
it is very young and hasn't experienced dynamical heating or radial
migration.

@@

\acknowledgements
It is a pleasure to thank
  Jeremy Magland (\acronym{SCDA}), and
  Eddie Schlafly (\acronym{LBNL})
for valuable discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)

>>>>

\begin{abstract}
% Context:
Chemical tagging promises to use surface chemical abundance measurements to find sets of stars
that were born together but which are now in spatially extended distributions.
This idea has not previously yielded scientific successes, probably because of the
noise and incompleteness in chemical-abundance measurements.
However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
delivering $\geq 15$ individual abundances for $\sim 10^5$ stars with precisions around 0.04~dex.
% Aims:
We test the chemical-tagging hypothesis by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot the associated stars in phase space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
% Results:
We find that clusters in abundance space are indeed clusters in phase space.
We recover some known clusters and find other interesting structures.
This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
the prospects for future data sets are very good.
\end{abstract}

@@

We have demonstrated for the first time that, in a large-scale
spectroscopic survey of stars, stars that appear similar in
chemical-abundance properties are often associated physically in
clusters or other structures.
The reverse is well established---that is, that stars that are
associated physically in clusters are similar in chemical-abundance
properties.
That is, although they often show chemical diversity, that chemical
diversity is small, and follows well-defined trends or a subspace in
abundance space (for example, \citealt{gratton, meszaros, bovy},
and references within those).
However, it was not known---prior to this work---whether abundance-space information would be
informative enough, or \emph{measurable} precisely enough to find this
structure in surveys dominated by a large mix of stars from different
origins and ages (and there was good reason to be pessimistic;
\citealt{ting}).
We have demonstrated with a straightforward (and in no way optimal)
clustering in chemical-abundance space that it is easy to find co-eval
stellar structures.
Although this is very strong evidence that chemical tagging will work,
we have not performed, in any sense, the canonical procedure of
looking for features that are consistent with being delta functions in
chemical-abundance space (convolved with a measurement noise
distribution); we have looked only for over-densities, not delta
functions.

What made our success here possible is substantial improvements in precision
delivered by \thecannon.
We don't have an absolute measure of the precision of the measurements,
but from looking at open and globular clusters, it appears to be on the
order of or better than 0.04~dex for the median element in the list of 15.
These improvements are discussed in detail in the companion papers
\citep{casey16, ness16} about the abundance measurements.
We believe these improvements come from a combination of factors, not
limited to (a)~improvements in the determination of pseudo-continuum,
(b)~the use of more spectral range than just unblended element windows
\citep{aspcap}, and (c)~accurate spectral predictions
(\thecannon\ delivers accurate predictions because it is fit to
observed data). 
Furthermore, because abundance space is high in dimensionality, even
small improvements in abundance measurement get taken to a significant
power when thinking about information gains.

@@

The k-means algorithm employed here is by no means optimal, and other clusters
we identify are accompanied by a few stars that are not currently co-spatial.
Those stars may be unassociated interlopers that have been misclassified,
or they may be true cluster members that are now unbound. However in the
case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
the situation is far more extreme. We identify a group of stars with
highly similar abundances in 15 dimensions that are now spread throughout
the Galaxy. Because some stars are candidate members of the massive globular
cluster M53, it provides tantalizing possibility that these stars may
indeed be all members of the same co-natal gas cloud, implying that 
these stars have been accreted onto the halo by a single globular
cluster early in the Milky Way's formation. While speculative,
this example demonstrates the promise of chemical tagging.

The abundance-space group of stars in the disk structure shown in
\figurename~\ref{fig:disk} may be associated with a very thin
Milky-Way bar component reported previously \citep{wegg}, which is
found to exist predominantly toward the end of the bar (at $l$
$\approx 30$~deg).
The stars in this structure are metal-rich and have a low
[C/N] ratio and are therefore likely young (following the [C/N]--mass
relationship; \citealt{martig}).
It is possible that this structure has very low scale height because
it is very young and hasn't experienced dynamical heating or radial
migration.

@@

\acknowledgements
It is a pleasure to thank
  Jeremy Magland (\acronym{SCDA})
for valuable discussions.
This project was funded in part by
  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
  \acronym{NASA} (grant \acronym{NNX12AI50G}),
  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
  the European Research Council under the
  European Union's Seventh Framework Programme (FP~7)
  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

@@

Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, 816, 10
